---
title: "session09 - exercises"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: false
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, lme4, brms, bayestestR)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Introduction

The main **learning goals** of this week's practical parts are:

-   getting an intuition about the gap between verbal hypotheses and formal hypotheses

# The data

In this dataset, we will be analyzing data from @kilpatrick2023random:

In a nutshell, the authors were interested in whether there is a relationship between the form of a word and its meaning by looking at Japanese Pokemon names.

Let's have a look at the data:

```{r}
#| label: data

# Load in the dataset and already remove irrelevant stuff:
pokemon <- read_csv('../data/Japanese_Pokemon.csv') |> 
  select(-`Sum of values`, -Checksum)

# check data set:
pokemon

```

What you see here are the following variables:

Variables:

`Designation`: source of Pokemon name. Official = the official names of Japanese Pokemon; Elicited = elicited names from naive participants.

`pokemonname`: Pokemon name in Japanese

`Output`: Pokemon name in transcribed form

`Length`: Amount of phones in name

`Evolution`: Evolution stage of Pokemons: 1 = pre-evolution; 2 = post-evolution, NA = do not evolve

`f` = amount of /f/ sounds in name

`T` = amount of /T/ sounds in name

...

and many more different sounds.

Now imagine researchers want to use this data set to *test* the following hypothesis:

*The evolution stage of a Pokemon is reflected in aspects of its Japanese name*

# Exercise

## A: Preregister your analysis

Take the hypothesis and what you know about the data set and its structure to 
predict a very specific relationship that you can test. You are allowed to look 
at the dataset, but do not look at concrete relationships that might be a 
manifestation of the hypothesis.

With iconicity research in mind (e.g. bigger things = longer words) and Pokemon 
knowledge (later stage of evolution = bigger and stronger), my analysis will 
focus on whether Pokemon names of later evolutions have longer words than 
Pokemon in early stages of evolution. I will only look at the official names 
since those show the original plan of the creators. I will remove NAs from 
the evolution column in my analysis so that I can look at the changes in 
evolution from level 1 to 2.

## B: Run your preregistered analysis

```{r, pokemon preregistered analysis}

#Doing some filtering for my analysis
poke_agg <- pokemon |>
  filter(Designation == "Official", !is.na(Evolution)) |>
  mutate(Evolution = as.factor(Evolution))

#Contrast coding
contrasts(poke_agg$Evolution) <- contr.sum(2)

#model
poke <- lm(Length ~ Evolution, data = poke_agg)

#Priors
get_prior(poke, poke_agg)

my_priors <- c(prior(normal(8, 3), class = "Intercept"),
              prior(normal(0, 20), class = "b"))

#Running the model
pokemod <- brm(Length ~ Evolution,
            family = gaussian(),
            prior = my_priors,
            data = poke_agg,
            chains = 4,
            cores = 4,
            iter = 2000,
            warmup = 1000,
            thin = 1,
            seed = 101)

#Cheking the output            
pokemod

#Double checking levels
levels(poke_agg$Evolution)
contrasts(poke_agg$Evolution)

p_direction(poke_agg)

```

My preregistered analysis showed that the hypothesis did not find support 
in the data. Pokemon's official name length does not reflect the evolution stage. 
I found that it is plausible that when the evolution stage of the pokemon goes 
from 1 to 2, the name would be 0.90 letters shorter, with a 95% credible 
interval of 1.12 letters shorter and 0.68 letters shorter. It is 100% sure 
that the direction of the effect is negative, that is, later evolution being 
smaller in name length.

## C: Show contradicting results

Now explore the data and find two alternative way to test the research 
hypothesis that yield convincing results that go in opposite directions 
based of your findings in (A). Justify all your choices verbally after the fact.

If I use NAs and 2 instead of 1 and 2 I get one more result that rejects the 
hypothesis. The reasoning for treating NA as 1 is that the way they put NAs in 
the dataset was strange and they were treating Pokemon with 3 evolution steps 
(like Pichu, Pikachu and Raichu) as 1, NA and 2. I will use both official 
and elicited names.

```{r, pokemon non-preregistered analysis}

#Doing some filtering for my analysis
poke_new <- pokemon |>
  mutate(Evolution = as.character(Evolution)) |>
  filter(Evolution != "1" | is.na(Evolution)) |>  
  mutate(
    Evolution = replace_na(Evolution, "1"),
    Evolution = as.factor(Evolution)
  )

#Contrast coding
contrasts(poke_new$Evolution) <- contr.sum(2)

#model
poken <- lm(Length ~ Evolution, data = poke_new)

#Priors
get_prior(poken, poke_new)

my_new_priors <- c(prior(normal(8, 3), class = "Intercept"),
              prior(normal(0, 20), class = "b"))

#Running the model
pokenew <- brm(Length ~ Evolution,
            family = gaussian(),
            prior = my_new_priors,
            data = poke_new,
            chains = 4,
            cores = 4,
            iter = 2000,
            warmup = 1000,
            thin = 1,
            seed = 101)

#Cheking the output            
pokenew

#Probability of direction
p_direction(poke_new)

```

By doing what I did in the treatment of the data, the change between lower 
and higher evolutions is very likely to be smaller (0.42), with a CI of -0.68 
to -0.16, but is still in the direction that "disproves" the hypothesis, and 
the probability of direction for a negative effect is 100%.

I will now do an analysis that will show that the evolution stage is reflected 
in the name of the pokemon. Iconicity research has shown that /i/ is associated 
with smaller things, so, lower evolutions should have more /i/s than higher 
evolutions. I don't use NAs and keep my analysis looking at 1 and 2 in the 
Evolution. I only look at elicited names since that will give the intuitions 
that people have spontaneously, which may tap better into iconicity than 
official names, which potentially go through several considerations 
and decision rounds.

```{r, pokemon non-preregistered analysis2}

#Doing some filtering for my analysis
poke_i <- pokemon |>
  filter(Designation == "Elicited") |>
  filter(!is.na(Evolution)) |>
  mutate(Evolution = as.factor(Evolution))

#Contrast coding
contrasts(poke_i$Evolution) <- contr.sum(2)

#model
pokeplo <- lm(i ~ Evolution, data = poke_i)

#Priors
get_prior(pokeplo, poke_i)

my_newest_priors <- c(prior(normal(1, 3), class = "Intercept"),
              prior(normal(0, 4), class = "b"))

#Running the model
pok <- brm(i ~ Evolution,
            family = gaussian(),
            prior = my_newest_priors,
            data = poke_i,
            chains = 4,
            cores = 4,
            iter = 2000,
            warmup = 1000,
            thin = 1,
            seed = 101)

#Cheking the output            
pok

#Probability of direction
p_direction(pok)

```

The model shows that it is likely that going from evolution 1 to 2, elicited 
pokemon names will have 0.04 less /i/, with a credible interval of -0.14 and 
0.06. So, 0 is still a possible number.

Looking at probability of direction, slightly over 78% of values are in the 
negative direction for the effect, so fewer /i/s the bigger a pokemon is. 
This means that the effect possibility is not very robust. There is still 
a good chance that the effect goes in the opposite direction.

## D: Researcher degrees of freedom

Note down all dimensions of researcher degrees of freedom that make the 
mapping of the hypothesis onto your data analysis vague. How many possible 
combinations can you think of?

The hypothesis does not specify what it means by evolution stage, and it does 
not specify at all what is meant by "will be reflected". Reflected how? 
Bigger pokemon having longer names? More voiced obstruents? Less vowels? 
More back vowels? Less high vowels? There are many possible combinations. 
Also, it does not specify at all how we're supposed to deal with the NAs, 
or how the official name and the elicited name are to play a role in the 
analysis. Also, we don't really know how the names were elicited so it was 
hard to decide on what model could best capture things. Did each informant 
provide 1 elicitation, or how many did they provide?

So, you can "easily p-hack" the results depending on 
which of all these paths you take.

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
